---
title: "Project 1"
author: "Dirk Hartog"
date: "2023-09-24"
output: pdf_document
---
# In this project, you’re given a text file with chess tournament results where the information has some structure. Your job is to create an R Markdown file that generates a .CSV file (that could for example be imported into a SQL database). 

```{r load libraries, echo = TRUE, message = FALSE}
library(tidyverse)
```


```{r read in the file}
url <- "https://raw.githubusercontent.com/D-hartog/DATA607/main/PROJECT1/chess.txt"

chess_txt <- read.csv("https://raw.githubusercontent.com/D-hartog/DATA607/main/PROJECT1/chess.txt")
head(chess_txt, 9)
```

# Separate out the rows with the information we need and create two columns using bind_cols and save it to a new data frame data.

```{r seperate and bind}
data <- bind_cols(chess_txt[seq(4,195,3),], chess_txt[seq(5,195,3),])
# Rename the two columns
colnames(data) <- c("top_row", "bottom_row")
head(data, 6)
```

# Use the seperate_wider_delim to separate each column in the top_row by "|" to create new columns with the player's name and total points.

```{r seperate top_row column, warning=FALSE}
data <- data %>%  
  separate_wider_delim(
    top_row,
    delim = "|",
    names = c("num", "player_name", "total_points"), 
    too_many = "debug"
  )
data[,c("num","top_row", "top_row_ok", "top_row_pieces")] <- NULL
```

# Repeat the above to the bottom_row column to get the state and pre-rank for each palyer.  Remove uneccessary columns.  
```{r separate bottome_row column, warning=FALSE}
data <- data %>%  
  separate_wider_delim(
    bottom_row,
    delim = "|",
    names = c("state", "pre_rank"), 
    too_many = "debug"
  )

data[, c("bottom_row","bottom_row_ok", "bottom_row_remainder", "bottom_row_pieces")] <- NULL

head(data)
```

# Extract the pre-rank for each player from the pre-rank column using a combination of a regular expression to parse the numbers and a for loop to grab the pre - rank number. 
```{r}
rank_match_list <- str_match_all(data$pre_rank, "\\d{3,4}")
rank_match_list[[1]]

#The player rank is the third number in each of the 64 lists
p_rank <- c()
for (i in 1:64){
  player_pr <- rank_match_list[[i]][3,1]
  p_rank <- append(p_rank, player_pr)
}

#Assign the p_rank vector back into the pre-rank column 
data$pre_rank <- p_rank 

head(data)
```

# Finding the average pre-ranks of each players opponents

```{r}
# Parse out the opponents from the "top_row_remainder" column
opponents_list <- str_match_all(data$top_row_remainder, "\\d{1,2}")

# Loop through the list of opponents and grab each pre-ranking to calculate the mean 
opp_pre_rank <- c()

for(i in c(1:64)){
  opponents <- opponents_list[[i]]
  opponents_pr <- data[opponents[1:length(opponents),1],"pre_rank"]
  opponents_pr_num <- as.numeric(unlist(opponents_pr))
  avg_opp_rank <- mean(opponents_pr_num)
  opp_pre_rank <- append(opp_pre_rank,avg_opp_rank)
}

# Round the vector with all the mean opponenet rankings and set it to a new column and drop the top_row_remainder
data$opponent_avgpr <- round(opp_pre_rank)
data$top_row_remainder <- NULL

head(data)
```

# Re-order the columns, clean them and convert the data types to numeric where appropriate
```{r}
data <- data[,c(1,3,2,4,5)]

data$player_name <- str_trim(data$player_name)
data$state <- str_trim(data$state)
data$total_points <- str_trim(data$total_points)
data$total_points <- as.numeric(data$total_points)
data$pre_rank <- as.numeric(data$pre_rank)

data
```

