---
title: 'Week 9: Web APIs'
author: "Dirk Hartog"
date: "2023-10-29"
output:
  html_document:
    df_print: paged
---
# The New York Times web site provides a rich set of APIs, as described here: https://developer.nytimes.com/apis. 

## Task: Choose one of the New York Times APIs, construct an interface in R to read in the JSON data, and transform it into an R DataFrame.

### Step 1: Load libraries 
```{r setup, include=FALSE}
library(httr2)
library(httr)
library(jsonlite)
library(tidyverse)
```

### Create a request to the top stories NY Times API
```{r API request}
req <- request("https://api.nytimes.com/svc/topstories/v2/science.json?api-key=4ZFLCUM2HxGjmPqm5PvB5InkBUcM7n2I")
```

```{r check aspects of the request}
req

req %>% req_headers("Accept" = "application/json")

req %>% req_body_json(list(x = 1, y = 2))
```
### Create a response to read data into a data frame
```{r response}
resp <- req_perform(req)
resp %>% resp_content_type()
resp <- resp %>% resp_body_json()
```

### Create a data frame
```{r data frame}
df <- tibble(resp)
df <- df[6,]
df <- df %>% unnest_longer(resp) %>%
  unnest_wider(resp)
head(df)
```

### Select rows and columns we are interested in and create a new df top_stories
```{r}
top_stories <- df %>% select(section, title, abstract, url, byline, published_date)
top_stories <- top_stories[c(-1,-2),]
glimpse(top_stories)
```

### Clean and tidy byline column
```{r }
# Extract author name and clean white space for easier tidying 
top_stories$byline <- str_trim(str_extract(top_stories$byline, "\\s\\D*"))

# Widen data frame into author first name and last name 
top_stories <- top_stories %>% separate_longer_delim(cols = byline, delim = " and ")
top_stories <- top_stories %>% separate_wider_delim(cols = byline, 
                            delim = " ", 
                            names = c("author_fname", "author_lname"),
                            too_many = "merge")
```

### Clean published date column
```{r clean published date}
top_stories$published_date <- as.Date(top_stories$published_date) 
top_stories$published_date <- ymd(top_stories$published_date)
```

### View final data frame
```{r}
head(top_stories)
```

### Data Visualization: The number of articles from each section of the news paper that make up the top stories 

```{r visualization}
top_stories %>% select(section, title) %>% distinct() %>% 
  ggplot(aes(x = section, fill = section)) + 
  geom_bar() +
  geom_text(aes(label = ..count..), stat = "count", vjust = 3, color = "white") +
  ggtitle("Top Stories By Section") + 
  xlab("Section") +
  theme(legend.position = "none",
        axis.title.x = element_text(color="black",size=10),
        axis.text.x = element_text(size=8),
        plot.title = element_text(color = "black",
                                  size=16))
```

