---
title: 'Wk 6 Data Transformation: Movies and Tv'
author: "Dirk Hartog"
date: "2023-10-08"
output:
  pdf_document: default
  html_document: default
---

```{r load library, include=FALSE}
library(tidyverse)
library(stringr)
```

## The first step was to read in the csv file from guthub.

```{r cars}
movieurl <- "https://raw.githubusercontent.com/D-hartog/DATA607/main/PROJECT2/movies_untidy.csv"
movies_tv <- read_csv(movieurl)
head(movies_tv)
summary(movies_tv)
```

## Before doing any transformations or analysis I wanted to clean up data and organize it a little differently. 

1. The first column I worked on was the Year column. Since I was interested in using the release year of a movie or TV show later in my analysis, I needed to extract the first year listed and clean up the string.

```{r year column, echo=FALSE}
movies_tv <- movies_tv %>% mutate(RELEASE_YEAR = str_extract(movies_tv$YEAR, "\\d{4}"))
movies_tv$RELEASE_YEAR <- as.numeric(movies_tv$RELEASE_YEAR)
```

2. This data set had both TV shows and Movies so I thought it would have been a good idea to try and label each observation accordingly. I created another column designating which observation was a TV show or a MOVIE based on certain criteria: 
- Whether or not the original YEAR value had more than one year or if it had a hypen indicating that the program was still running. 
- TV shows are usually 30 or 60 minutes long. I used the RunTime columnm to find those observations that were less than 75 minutes. I included observations longer than 60 since some TV shows do have an occasional special longer episode. (I understand that this may not have been the most accurate as this cut off was based on personal experience as a mediocore TV watcher and reading some articles on the internet).

```{r create col TYPE}
movies_tv <- movies_tv %>% 
  mutate(TYPE=(ifelse (str_detect(movies_tv$YEAR, "\\(\\d{4}.{2,}") | RunTime < 75, "TV", "MOVIE")))

head(movies_tv, 3)
```

3a. There was a lot of cleaning to do in the STARS column as it listed both the actors and director in one cell. I extracted the actors from the STARS column and created a new column called ACTORS. I did the same thing for the directors listed and created a new column called DIRECTOR.
```{r}
# Find and extract the string listing all the actors and assign it to a new column. Trim any white space or new line characters
movies_tv <- movies_tv %>% 
  mutate(ACTORS = str_extract(movies_tv$STARS, "Stars:(\\n.*)+"))
movies_tv$ACTORS <- str_trim(movies_tv$ACTORS)

# I could not figure out how to clean and extract exactly what I want in one step so it was a multi-step iteration. The steps above were repeated but extracting only the names of the actors

movies_tv <- movies_tv %>% 
  mutate(ACTORS = str_extract(movies_tv$ACTORS, "(\\n.*)+"))
movies_tv$ACTORS <- str_trim(movies_tv$ACTORS)
```

3b. At this point it made sense to transform the ACTORS column so that each actor was it's own observation transforming the data frame into a longer one. 
```{r seperate string of ACTORS }
movies_tv <- movies_tv %>% separate_longer_delim(ACTORS, delim = ", \n")
```

3c. As described above the director was extracted in a similar way from the original STARS column to create a new column called DIRECTOR.
```{r}
movies_tv <- movies_tv %>%
  mutate(DIRECTOR = str_extract(movies_tv$STARS, "Director:.*\\n.*\\n"))

movies_tv <- movies_tv %>%
  mutate(DIRECTOR = str_extract(movies_tv$DIRECTOR, "\\n.*"))

movies_tv$DIRECTOR <- str_trim(movies_tv$DIRECTOR)
```

4. In order to do some analysis on the genres of programming this data set contained, a bit of cleanging and transforming need to be done. I extracted the first listed genre into the GENRE column and kept the others listed in another column.
```{r}
movies_tv$GENRE <- str_trim(movies_tv$GENRE)

movies_tv <- movies_tv %>% mutate(GENRE = str_extract(movies_tv$GENRE, "(^[A-Z][a-z]*)"), GENRE_OTHER = str_extract(movies_tv$GENRE, "\\s([A-Z][a-z]*),\\s([A-Z][a-z]*)"))

movies_tv$GENRE_OTHER <- str_trim(movies_tv$GENRE_OTHER)

```

5. I wanted to change some column names and select only the columns I needed for analysis. 
```{r}
movies_tv <- movies_tv %>% rename(RUN_TIME = RunTime) %>% 
  select(MOVIES, GENRE, GENRE_OTHER, RATING, VOTES, RUN_TIME, RELEASE_YEAR, TYPE,
         ACTORS, DIRECTOR) 

head(movies_tv)

# Save tidy data frame to csv
write.csv(movies_tv,file='/Users/dirkhartog/Desktop/CUNY_MSDS/DATA_607/PROJECT2/movies/movies_tv.csv', row.names=FALSE)
```

## DATA ANALYSIS :  The data was cleaned and transformed in order to investigate the relationships between the ratings of tv shows and/or movies and genre and actor in the data. 

1a. Finding the top 15 genres with the highest average rating 
```{r}
top15avg <- movies_tv %>% filter(TYPE == "MOVIE") %>%
  drop_na(GENRE) %>%
  group_by(GENRE) %>%
  summarise(Average = round(mean(RATING, na.rm = TRUE),1),
            Max = max(RATING, na.rm = TRUE),
            Min = min(RATING, na.rm = TRUE)) %>%
  arrange(desc(Average)) %>%
  head(15)

top15avg
```

1b. Visualize the distribution of the ratings among the top 15
 
```{r genre rating distribution}
#create a new data frame with only the top15 highest rated genres
df <- movies_tv %>% drop_na(GENRE) %>%
  filter(GENRE %in% top15avg$GENRE)

ggplot(data = df, mapping = aes(y = RATING,x = GENRE)) +
  geom_boxplot() +
  ggtitle("15 Highest Avg. Rated Genres") +
  ylab("Ratings") +
  xlab("Genres") +
  theme(axis.title.x = element_text(color="darkgreen",size=12),
        axis.title.y = element_text(color="red", size=12),
        axis.text.x = element_text(size=10),
        axis.text.y = element_text(size=10),
        plot.title = element_text(color="darkblue",
                            size=18)) +
  coord_flip()
```

**Conclusions** 
Using box plots is helpful to visualize a lot of data among a variable and making comparisons across variables with categorical data types. In this plot we can see the distribution of ratings within the top 15 movie genres with the highest average rating. Here we can see that the median values of each genre are between 6 and 7.5. This also gives us an idea of the range of values in each genre and any outlines present. It looks like there are few and maybe even 1 value in the Film and History genres which future considerations may be looking at the most common genres listed.

2a. Find the top 10 actors or actresses who appeared the most in this data set
```{r}
movies_only <- movies_tv %>% filter(TYPE == "MOVIE") %>% 
  drop_na(ACTORS)

top10actors <- movies_only %>% group_by(ACTORS) %>%
  summarise(n = n()) %>%
  arrange(desc(n)) %>% 
  head(10)

top10actors
```

2b. Find the average rating of the movies the actors above were in. 
```{r}
df2 <- movies_only %>%filter(ACTORS %in% top10actors$ACTORS)


top10avg <- df2 %>% group_by(ACTORS) %>%
  summarise(Average = round(mean(RATING, na.rm = TRUE),1),
            Max = max(RATING, na.rm = TRUE),
            Min = min(RATING, na.rm = TRUE)) %>%
  arrange(desc(Average))

top10avg
```

2c. We want to evaluate this against the average of all movies in the data set
```{r}
top10avg <- top10avg %>% 
  mutate(Above_avg = ifelse(top10avg$Average > mean(movies_only$RATING, na.rm = TRUE), "YES", "NO"))
```

```{r}
ggplot(top10avg, aes(x = ACTORS, y = Average, fill = Above_avg)) +
  geom_col() +
  ggtitle("Avg Rating of the Top 10 Most Frequent Actors") +
  ylab("Ratings") +
  xlab("Actors") +
  theme(axis.title.x = element_text(color="darkgreen",size=12),
        axis.title.y = element_text(color="red", size=12),
        axis.text.x = element_text(size=10, angle = 80, hjust = 0.45, vjust = 0.5),
        axis.text.y = element_text(size=10),
        plot.title = element_text(color="darkblue",
                                  size=18))
```

**Conclusions**
In this plot we can see the average movie rating of each of the top 10 most most frequently listed actors in this data set. We can see that over half of the actors in this list reached above the average rating for all movies 'r mean(movies_only$RATING, na.rm = TRUE)' in the data set. I don't think we can draw strong conclusions about the influence an actor has on the raings of movies just with this plot alone. It might be interesting to expand this and look at the average movies ratings across all actors in the data set. 

