---
title: 'Wk 7: Working with XML, HTML and JSON in R'
author: "Dirk Hartog"
date: "2023-10-11"
output: pdf_document
---
```{r message=FALSE}
library(tidyverse)
library(jsonlite)
library(XML)
library(xml2)
library(stringr)
library(rvest)
```

#Pick three of your favorite books on one of your favorite subjects. At least one of the books should have more than one author. For each book, include the title, authors, and two or three other attributes that you find interesting.

## Step 1: Create three files which store the book’s information in HTML (using an html table), XML, and JSON formats and read the information form the web (in this case Github). 

**JSON file**
```{r read JSON}
# libraries used: jsonlite, dplyr, stringr 

json_url <- "https://raw.githubusercontent.com/D-hartog/DATA607/main/WK7/books.json"
json_file <- list(read_json(json_url))
books_js <- tibble(books = json_file)

# Unnest the list object from the json file

books_js <- books_js %>% unnest_wider(books) %>%
  unnest_longer(books) %>%
  unnest_wider(books)

# Expand the cell with a list of two authors 

# -- Unlist that cell into a vector
authors <- unlist(books_js$author_s[2])

# -- flatten the vector to create a string of the two authors and re-assign it into the cell
books_js$author_s[2] <- str_flatten(authors, ",")

# -- Use seperate rows function to separate the rows into two (one with each author)
books_js <- books_js %>% separate_rows(author_s, sep = ",")

# -- Clean up column names and values
colnames(books_js) <- books_js %>% names() %>% str_to_upper()

books_js$AUTHOR_S <- str_trim(books_js$AUTHOR_S)

books_js
```

**HTML file**
```{r read HTML}
# libraries used: rvest, dplyr, stringr

html_url <- "https://raw.githubusercontent.com/D-hartog/DATA607/main/WK7/books.html"

# read in the file form github
html_file <- read_html(html_url)

books_html <- html_file %>%
  html_table(fill = TRUE)

books_html <- books_html[[1]]

# Use seperate rows function to separate the rows into two (one with each author)
books_html <- books_html %>% separate_rows(author_s, sep = ",")

# Clean up column names and values
colnames(books_html) <- books_html %>% names() %>% str_to_upper()

books_html$AUTHOR_S <- str_trim(books_html$AUTHOR_S)

books_html
```

**XML file**
```{r read xml}
# libraries used: xml2, dplyr, stringr

xml_url <- "https://raw.githubusercontent.com/D-hartog/DATA607/main/WK7/books.xml"

# Read in xml file from url 
xml_file <- read_xml(xml_url)

# Extract all the book nodes, transform them into a list and bind_rows()
books_xml <- xml_file %>%
  xml_find_all(".//book") %>%
  as_list() %>% 
  bind_rows()

# Use seperate rows function to separate the rows into two (one with each author)
books_xml <- books_xml %>% separate_rows(author_s, sep = ",")

# Clean up column names and values
colnames(books_xml) <- books_xml %>% names() %>% str_to_upper()

books_xml$AUTHOR_S <- str_trim(books_xml$AUTHOR_S)

books_xml
```

For the xml and html files one the data frame was constructed, they did both look similar. When initally constructing the dataframe with the json file, the book with two authors were contain in a list within the author column. An extra step was required to unnest that list. Now all the data frames look the same. 
