---
title: 'Wk 6 Data Transformation: World Population'
author: "Dirk Hartog"
date: "2023-10-08"
output:
  html_document: default
  pdf_document: default
---
```{r message=TRUE}
library(tidyverse)
```

## We will looking at a data set containing global population counts for 234 countries or territories. 

## Read in the untidy .csv file from github 
```{r }
url <- "https://raw.githubusercontent.com/D-hartog/DATA607/main/PROJECT2/worldpop_untidy.csv"

world_pop <- read_csv(url)
head(world_pop)
glimpse(world_pop)
```

## CLEANING THE DATA

1. First I want to change the column names for later transformtion of the year columns .
```{r chang col names}
colnames(world_pop)[c(1:17)] <- c("RANK","CCAS", "COUNTRY_TERR", "CAPITAL","CONTINENT", "2022", "2020", "2015", "2010", "2000", "1990", "1980", "1970","AREA", "DENSITY", "GROWTH_RATE", "WORLD_POP_PCT")
```

## TIDY/TRANSFORMING THE DATA: For the analysis I want to compare growth rates by year as well as among continents. In order to make do this analysis we will need to created an observation for each country and year. 

1. I don't think that there is much transformation that needs to be done to the data execpt taking the year columns pivoting those columns to rows. 

```{r Transform year columns}
world_long <- world_pop %>%
  pivot_longer(
    cols = "2022":"1970",
    names_to = "YEAR",
    values_to = "POPULATION"
  )

glimpse(world_long)

write.csv(world_long,file='/Users/dirkhartog/Desktop/CUNY_MSDS/DATA_607/PROJECT2/world_pop/worldpop_tidy.csv', row.names=FALSE)
```

## DATA ANALYSIS AND VISUALIZATIONS 

1. Finding the total, min and max populations for each year and each and continent
```{r total populations in 2022}
world_long %>% 
  group_by(CONTINENT, YEAR) %>%
  summarize(TOTAL = sum(POPULATION, na.rm = TRUE),
         MAX = max(POPULATION, na.rm = TRUE),
         MIN = min(POPULATION, na.rm = TRUE))
```

```{r plot of continental population data}
world_long %>% group_by(CONTINENT, YEAR) %>% 
  summarise(TOTAL = sum(POPULATION, na.rm = TRUE)) %>%
  ggplot(aes(x = YEAR, y = TOTAL)) +
  geom_point(aes(color = CONTINENT)) +
  ggtitle("50 Year Trend in Continental Populations")
```

2. We can now pull out information for a given year and compare attributes, in this case among continents. Looking at some statistics and trends in growth rates from Asian and African countries in 2022

```{r}
world_long %>%  filter(YEAR == "2022") %>%
  group_by(CONTINENT) %>%
  summarise(Average_gr = mean(GROWTH_RATE, na.rm = TRUE))
```

```{r histogram for asian and african countries}
world_long %>% 
  filter(YEAR == "2022" & CONTINENT %in% c("Asia", "Africa")) %>%
  ggplot(aes(x = GROWTH_RATE)) +
  geom_histogram(bins = 20, aes(fill = CONTINENT)) +
  ggtitle("Histogram of Growth Rates Among Asian 
  and African Countries/Territories") +
  ylab("Count") +
  xlab("Growth Rate") + 
  theme(axis.title.x = element_text(color="darkgreen",size=15),
        axis.title.y = element_text(color="red", size=15),
        axis.text.x = element_text(size=10),
        axis.text.y = element_text(size=10),
        plot.title = element_text(color="darkblue",
                                  size=18))
```

3. Looking at any relationships in area size less than 5 million (km^2) and growth rate
```{r}
world_long %>% filter(YEAR == "2022" & AREA < 5000000 & CONTINENT %in% c("Asia", "Africa")) %>%
  ggplot(aes(x = AREA, y = GROWTH_RATE, color = CONTINENT)) +
  geom_point() +
  ggtitle("Area and Growth Rate of Asian 
  and African Countries/Territories in 2022") +
  ylab("Growth Rate") +
  xlab("Area (km^s)") + 
  theme(axis.title.x = element_text(color="darkgreen",size=15),
        axis.title.y = element_text(color="red", size=15),
        axis.text.x = element_text(size=10),
        axis.text.y = element_text(size=10),
        plot.title = element_text(color="darkblue",
                                  size=18))
```
## CONCLUSIONS
It is clear that Asian and African countries have seen a larger trend in the growth of their populations over the past 50 years. Despite this it does seem that currently average growth rates across the globe are pretty similar between 1.002 - 1.02. This growth rate might seem small but when talking about populations, a 1% grwoth rate is still a lot of people!
