---
title: "Healthy Aging"
author: "Dirk Hartog"
date: "2023-12-08"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(httr2)
library(httr)
library(jsonlite)
library(tidyverse)
library(utils)
library(pdftools)
library(maps)
library(mapproj)
library(purrr)
```

# INTRODUCION: 

Despite a slight drop in life expectancy (LE) in the United States from 2020 - 2021, estimates for the average LE in 2023 support the return to a pre pandemic level of 79.11 years. As life expectancy continues to rise we might also expect that our aging population will continue to increase as well. As people get older, their risk of chronic diseases increases such as dementia, heart disease, type 2 diabetes, arthritis, and cancers. 

# QUESTIONS TO INVESTIGATE: 

+ What are the trends in the population of individuals 65 years or older in the United States?

+ What are the leading causes of death for individuals aged 65 years or older in the United States. 

+ What are the trends in modifiable behaviors or risk factors that might be related to healthy aging? 

+ Is there any correlation between subjective cognitive decline and Alzheimer’s death rates in the united states?


# POPULATION TRENDS

#### Read/clean/tidy data

```{r}
# Clean and Tidy
url <- "https://raw.githubusercontent.com/D-hartog/DATA607/main/FINAL_PROJ/population_trend.csv"
population <- read_csv(url)

population <- population[1:52,]

head(population)

# Multiple variables stored in column names
population <- pivot_longer(population,
             cols = `2008__Children 0-18`:`2022__Total`,
             names_to = "Year_AgeGroup",
             values_to = "Count")

population$Year_AgeGroup <- str_replace_all(population$Year_AgeGroup, "__", ",")

population <- separate_wider_delim(population,
                     Year_AgeGroup,
                     delim = ",",
                     names = c("Year", "Group")
                     )
population$Footnotes <- NULL
population$Year <- as.numeric(population$Year)
head(population)
```

#### Visualize data

**1. Time series plot of us pop**

```{r}
# 1. Time series plot of us pop
population %>% filter(Group == "Total" & Location == "United States") %>%
  select(Year, Count) %>% 
  ggplot(aes(x = Year, y = Count /1000000)) + 
  geom_point(color = "red", shape = 25, size = 5) +
  geom_line(color = "darkblue", linetype = 1, linewidth = 1) +
  ggtitle("Population Growth (2008-2022)") + 
  xlab("Year") +
  ylab("Count (millions)") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.title.x = element_text(color="black",size=12),
        axis.title.y = element_text(color="black", size=12),
        axis.text.x = element_text(size=10),
        axis.text.y = element_text(size=10),
        plot.title = element_text(color = "black", size=22))
   
population %>% filter(Group == "Total" & Location == "United States") %>%
  select(Year, Count) %>% mutate(Pecent_change = Count)
```

**2. Proportion of population in each state**
```{r}
states <- map_data("state")
states$region <- str_to_title(states$region)

states <- states[order(states$order),]

# Find population in 2022
#population %>% filter(Location == "United States" & Year == 2022 & Group == "Total")

# State proportions of 65+ 

population %>% filter(Location != "United States" & Year == 2022 & Group == "65+") %>% 
  merge(states, by.x = "Location", by.y = "region") %>% 
  mutate(Percent = Count /324486000) %>% 
  arrange(group, order) %>% 
  ggplot(aes(x = long, y = lat)) + 
  geom_polygon(aes(group = group, fill = Percent), color = "white", linewidth = 0.1) + 
  coord_map() +
  scale_fill_gradient(low = "blue", high = "red") +
  theme_minimal() + 
  ggtitle("Proportion of  65+ By State (2022)") + 
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        plot.title = element_text(color = "black", size=24))

```

**3. Trends in the proportions of the population**

```{r}

# 1. Time series plot of us population by age group
population %>% filter(Location == "United States") %>%
  group_by(Group) %>% 
  ggplot(aes(x = Year, y = Count /1000000, color = Group)) + 
  geom_point(size = 0.75) +
  geom_line(linewidth = 0.75) +
  scale_y_log10() +
  ggtitle("Trends in US Population by Age Group") + 
  xlab("Year") +
  ylab("Population (millions)") +
  theme_minimal() + 
  theme(legend.position = "right",
        axis.title.x = element_text(color="black",size=10),
        axis.title.y = element_text(color="black", size=10),
        axis.text.x = element_text(size=6),
        axis.text.y = element_text(size=6),
        plot.title = element_text(color = "black", size=20))
 
# 2. Proportion of the us population by age
population %>% filter(Year == 2022 & 
                        Location == "United States") %>%
  mutate(Percent = Count /max(Count)) %>% 
    filter(Group != "Total") %>% 
  ggplot(aes(x = fct_rev(fct_reorder(Group,Percent)), y = Percent)) + 
  geom_col(aes(fill = Group)) +
  ggtitle("Population Proportion by Age (2022)") + 
  xlab("Age Group") +
  ylab("Percent") +
  theme_minimal() + 
  theme(legend.position = "none",
        axis.title.x = element_text(color="black",size=10),
        axis.title.y = element_text(color="black", size=10),
        axis.text.x = element_text(size=6),
        axis.text.y = element_text(size=6),
        plot.title = element_text(color = "black", size=20))
```


# LEADING CAUSES OF DEATH IN THE USA

### Read/clean/tidy data 

```{r}
# Read in pdf file from local directory 
## ISSUE - could not figure out how to load a pdf from github into R

# link: https://github.com/D-hartog/DATA607/blob/main/FINAL_PROJ/leading_causes1819.pdf
pdf1 <- c("leading_causes1819.pdf")
raw_text <- purrr::map(pdf1, pdf_text)

#Split pages 
raw_text <- purrr::map(raw_text, ~ str_split(.x,"\\n") %>% unlist)

#Concatenate the split pages
raw_text <- reduce(raw_text, c)

#Specify the start and end of the table data
table_start <- str_which((raw_text), "Data Brief 395: Mortality in the United States, 2019")
table_end <- str_which((raw_text), "… Category not applicable.")
table_end <- table_end[min(which(table_end > table_start))]

#Build the table and remove special characters 
table <- raw_text[(table_start):(table_end)]
table <- str_replace_all(table, "\\s{2,}", "|")
table <- table[8:21]
table[1] <- "Rank1|Cause of death (based on International Classification of Diseases,10th Revision [ICD–10])|Number_18|Percent_18|Rate_18|Number_19|Percent_19|Rate_19"
table[3] <- "AC|All causes|NA|2,839,205|100.0|723.6|2,854,838|100.0|715.2"
table[14] <- "AOC|All other causes|(residual)|744,312|26.2|…|758,167|26.6|…"
text_con <- textConnection(table)
data_table <- read.csv(text_con, sep = "|")

#Create a list of column names
colnames(data_table) <- c("Cause_of_death",
                          "ICD–10",
                          "Number_18",
                          "Percent_18",
                          "Rate_18",
                          "Number_19",
                          "Percent_19",
                          "Rate_19")
data_table <- tibble(data_table)

df <- data_table

df$`ICD–10` <- NULL

# Matching the df to what our other mortality data has 
df[df == "Diseases of heart"] <- "Heart disease"
df[df == "Malignant neoplasms" ] <- "Cancer"
df[df == "Accidents (unintentional injuries)"] <- "Unintentional injuries"
df[df == "Intentional self-harm (suicide)"] <- "Suicide"
df[df == "Cerebrovascular diseases"] <- "Stroke"
df[df == "Chronic lower respiratory diseases"] <- "CLRD" 
df[df == "Diabetes mellitus"] <-  "Diabetes"
df[df == "Nephritis, nephrotic syndrome and nephrosis"] <-  "Kidney disease"
df[df == "Alzheimer disease"] <-  "Alzheimer's disease"

df <- df %>% filter(Cause_of_death != "All other causes")

df$Rate_18 <- as.double(df$Rate_18)
df$Rate_19 <- as.double(df$Rate_19)

# Clean numbers column and convert the data type
df$Number_18 <- as.double(str_remove_all(df$Number_18, ","))
df$Number_19 <- as.double(str_remove_all(df$Number_19, ","))

df$Percent_18 <-  df$Percent_18 + 0.0
df$Percent_19 <- df$Percent_19 + 0.0

yr_18 <- df[,c(1,2,3,4)]
yr_18$Year <- 2018
colnames(yr_18) <- c("Cause_of_death", "Number", "Percent", "Rate", "Year")

yr_19 <- df[,c(1,5,6,7)]
yr_19$Year <- 2019
colnames(yr_19) <- c("Cause_of_death", "Number", "Percent", "Rate", "Year")

df <- rbind(yr_18, yr_19)
df <- df %>% select(Year, Cause_of_death, Number, Percent, Rate)
df$Percent <- NULL

```

```{r}

# LEADING CAUSES OF DEATH 2020 - 2021 PDF
# https://github.com/D-hartog/DATA607/blob/2bb1ad36a7459bbf61e5c9d6baff1f32638af115/FINAL_PROJ/leading_causes2021.pdf
pdf2 <- c("leading_causes2021.pdf")
raw_text2 <- purrr::map(pdf2, pdf_text)
# https://www.youtube.com/watch?v=bJH-S2iaxNo

#Split pages 
raw_text2 <- purrr::map(raw_text2, ~ str_split(.x,"\\n") %>% unlist)
#Concatenate the split pages
raw_text2 <- reduce(raw_text2, c)

#Specify the start and end of the table data
table_start2 <- str_which((raw_text2), "Data Brief 456. Mortality in the United States, 2021")
table_end2 <- str_which((raw_text2), "… Category not applicable\b")
table_end2 <- table_end2[min(which(table_end2 > table_start2))]

#Build the table and remove special characters 
table2 <- raw_text2[(table_start2):(table_end2)]
table2 <- str_remove_all(table2, "\b")
table2 <- str_replace_all(table2, "\\s{2,}", "|")

table2 <- table2[8:24]

table2[1] <- "Rank1|Cause_of_death|Number_20|Percent_20|Rate_20|Number_21|Percent_21|Rate_21"
table2[3] <- "AC|All causes|ICD_10|3,383,729|1000|8354|3,464,231|1000|8797"
table2[7] <- "4|Accidents (unintentional injuries)|(V01–X59,Y85–Y86)|200,955|59|576|224,935|65|647" 
table2[13] <- "9|Chronic liver disease and cirrhosis|(K70,K73–K74)|51,642|15|133|56,585|16|145"
table2[15] <- "10|Nephritis, nephrotic syndrome and nephrosis (kidney disease)|(N00–N07,N17–N19,N25–N27)|52,547|16|127|54,358|16|136"
table2[17] <- "AOC|All other causes|(residual)|879,091|260|…|882,775|255|…"
table2 <- table2[c(1,3,4,5,6,7,9,10,11,12,13,15,17)]
text_con2 <- textConnection(table2)
data_table2 <- read.csv(text_con2, sep = "|")

#Create a list of column names
colnames(data_table2) <- c("Cause_of_death",
                          "ICD–10",
                          "Number_20",
                          "Percent_20",
                          "Rate_20",
                          "Number_21",
                          "Percent_21",
                          "Rate_21")
data_table2 <- tibble(data_table2)

df2 <- data_table2

df2$`ICD–10` <- NULL

df2$Rate_20 <- as.double(df2$Rate_20)
df2$Rate_21 <- as.double(df2$Rate_21)

# Clean numbers column anc convert data type
df2$Number_20 <- as.double(str_remove_all(df2$Number_20, ","))
df2$Number_21 <- as.double(str_remove_all(df2$Number_21, ","))

df2$Percent_20 <-  df2$Percent_20 + 0.0
df2$Percent_21 <- df2$Percent_21 + 0.0

yr_20 <- df2[,c(1,2,3,4)]
yr_20$Year <- 2020
colnames(yr_20) <- c("Cause_of_death", "Number", "Percent", "Rate", "Year")

yr_21 <- df2[,c(1,5,6,7)]
yr_21$Year <- 2021
colnames(yr_21) <- c("Cause_of_death", "Number", "Percent", "Rate", "Year")

df2 <- rbind(yr_20, yr_21)
df2 <- df2 %>% select(Year, Cause_of_death, Number, Percent, Rate)
df2$Percent <- NULL

df2 %>% group_by(Year) %>% arrange(Year,desc(Rate))

df2[df2 == "Diseases of heart"] <- "Heart disease"
df2[df2 == "Malignant neoplasms (cancer)"] <- "Cancer"
df2[df2 == "Accidents (unintentional injuries)"] <- "Unintentional injuries"
df2[df2 == "Intentional self-harm (suicide)"] <- "Suicide"
df2[df2 == "Cerebrovascular diseases (stroke)"] <- "Stroke"
df2[df2 == "Chronic lower respiratory diseases"] <- "CLRD" 
df2[df2 == "Diabetes mellitus"] <-  "Diabetes"
df2[df2 == "Nephritis, nephrotic syndrome and nephrosis (kidney disease)"] <-  "Kidney disease"
df2[df2 == "Alzheimer disease"] <-  "Alzheimer's disease"

df2 <- df2 %>% filter(Cause_of_death != "All other causes")


df2$Rate <- df2$Rate /10
```

### Visualizations 

```{r}
# Combine data-frames into one
# ISSUE - could not figure out how to load a zipped from github into R

leading_mortality <- read_csv("https://raw.githubusercontent.com/D-hartog/DATA607/main/FINAL_PROJ/LeadingCausesofDeath.csv")

leading_mortality <- leading_mortality %>% filter(State == "United States") %>% 
  select(Year, `Cause Name`, Deaths, `Age-adjusted Death Rate`)

colnames(leading_mortality) <- c("Year", "Cause_of_death", "Number", "Rate")

leading_mortality <- rbind(df, df2, leading_mortality)

head(leading_mortality)

leading_mortality %>% filter(Cause_of_death != "All causes") %>%
  group_by(Year) %>% arrange(Year,desc(Rate)) %>% 
  ggplot(aes(x = Year, y = Rate, color = Cause_of_death)) + 
  geom_point(size = 0.5) + 
  geom_line() + 
  scale_y_log10() + 
  ggtitle("Trends in the Leading Causes of Death in the US") + 
  xlab("Year") +
  ylab("Rate (deaths/100,000)") +
  theme_minimal() + 
  theme(axis.title.x = element_text(color="black",size=10),
        axis.title.y = element_text(color="black", size=10),
        axis.text.x = element_text(size=6),
        axis.text.y = element_text(size=6),
        plot.title = element_text(color = "black", size=20))

leading_mortality %>% filter(Cause_of_death == "All causes") %>%
  ggplot(aes(x = Year, y = Rate)) +
  geom_line() +
  ggtitle("Trends in the US Death rate") + 
  xlab("Year") +
  ylab("Rate (deaths/100,000)") +
  theme_minimal() + 
  theme(legend.position = "none",
        axis.title.x = element_text(color="black",size=10),
        axis.title.y = element_text(color="black", size=10),
        axis.text.x = element_text(size=6),
        axis.text.y = element_text(size=6),
        plot.title = element_text(color = "black", size=20))

```
```{r}
# Percent change in select diseases 
alz_pct_change <- ((31.0 - 16.5)/16.5) * 100
heart_pct_change <- ((173.8 - 266.5)/266.5) * 100
cancer_pct_change <- ((146.6 - 200.8)/200.8) * 100
uninten_pct_change <-((64.7 - 35.3)/35.3) * 100
tibble(Alzheimers = alz_pct_change,Heart_desiease = heart_pct_change, Caner = cancer_pct_change, Unintentional = uninten_pct_change)
```

# CAUSES OF DEATH IN 65+ 

### Read/clean/tidy data 

```{r top 10 leading causes}
# read in data using - API

#https://data.cdc.gov/NCHS/NCHS-VSRR-Quarterly-provisional-estimates-for-sele/489q-934x/about_data
req <- request("https://data.cdc.gov/resource/489q-934x.json")

resp <- req_perform(req)

resp <- resp %>% resp_body_json()

# tidy data - transform to wider table 

mortality65_plus <- unnest_wider(tibble(resp), resp)

# separate year and quarter to make it easier to filter data 

mortality65_plus <- mortality65_plus %>% separate_wider_regex(year_and_quarter,
                                   c(year = "\\d{4}", quarter = "\\sQ\\d"))

# clean columns and convert data types 
mortality65_plus$rate_overall <- as.double(mortality65_plus$rate_overall)
mortality65_plus$quarter <- str_trim(mortality65_plus$quarter)

# filter age groups, rate_type, causes, quarter and time period 
# pivot age columns longer to construct visuals 

mortality65_plus <- mortality65_plus %>% select(year, quarter, time_period, cause_of_death,rate_type, unit, rate_overall,rate_65_74, rate_age_75_84,rate_age_85_plus) %>% 
  filter(rate_type == "Crude" & cause_of_death != "All causes" & quarter == "Q4" &
           time_period == "12 months ending with quarter") %>%
  pivot_longer(cols = rate_65_74:rate_age_85_plus,
               names_to = "group", 
               values_to = "rate")

mortality65_plus$rate <- as.double(mortality65_plus$rate)

head(mortality65_plus)

mortality65_plus %>% group_by(group, year) %>% top_n(10, rate) %>% 
  ggplot(aes(x = year, y = rate, color = cause_of_death)) + 
  geom_line(aes(group = cause_of_death)) +
  facet_grid(~group) +
  scale_y_log10() +
  theme(legend.position = "bottom") +
  guides(color = guide_legend(ncol = 4)) +
  ggtitle("One Year Trend: Top 10 Causes of Death Per Age Group") + 
  xlab("Year") +
  ylab("Rate (deaths/100,000)") +
  theme_minimal() + 
  theme(legend.position = "bottom",
        axis.title.x = element_text(color="black",size=10),
        axis.title.y = element_text(color="black", size=10),
        axis.text.x = element_text(size=10),
        axis.text.y = element_text(size=10),
        plot.title = element_text(color = "black", size=16))
  

mortality65_plus %>% filter(year == 2021) %>% group_by(group) %>%
  top_n(10, rate) %>% 
  ggplot(aes(x = fct_rev(fct_reorder(cause_of_death,rate)), y = rate, fill = cause_of_death)) +
  geom_col() +
  facet_grid(~group) + 
  scale_y_log10() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  ggtitle("Top 10 Leading Causes of Death in Ages 65+ (2021)") + 
  xlab("Cause of death") +
  ylab("Rate (deaths/100,000)") +
  theme_minimal() + 
  theme(legend.position = "none",
        axis.title.x = element_text(color="black",size=10),
        axis.title.y = element_text(color="black", size=10),
        axis.text.x = element_text(size=8, angle = 60, vjust = 1, hjust = 1),
        axis.text.y = element_text(size=10),
        plot.title = element_text(color = "black", size=12))

```


# HEALTH INDICATORS: BEHAVIORAL RISK FACTOR SURVEILANCE SYSTEM (BRFSS)

```{r}
brfss <- read_csv("Alzheimer_s_Disease_and_Healthy_Aging_Data_20231115.csv")

glimpse(brfss)

brfss <- brfss %>% select(-c("Response", "Sample_Size", "StratificationCategory3", "Stratification3",
                        "ResponseID", "StratificationCategoryID3", "StratificationID3",
                        "Report", "Data_Value_Unit", "DataValueTypeID", "Data_Value_Footnote_Symbol",
                        "Datasource", "StratificationCategory1", "StratificationCategoryID1",
                        "StratificationCategoryID2", "RowId"))


brfss <- pivot_wider(brfss, 
            names_from = "StratificationCategory2",
            values_from = "Stratification2")

brfss$`NA` <- NULL

# Filter out Gaum and Puerto Rico
brfss <- brfss %>% filter(LocationDesc != "Guam" & LocationDesc != "Puerto Rico")


# Clean Gender and Race columns
brfss$Gender <- as.character(brfss$Gender)
brfss$Gender[brfss$Gender == "NULL"] <- NA

brfss <- brfss %>% rename("Race_Ethnicity" = `Race/Ethnicity`)
brfss$Race_Ethnicity <- as.character(brfss$Race_Ethnicity)
brfss$Race_Ethnicity[brfss$Race_Ethnicity == "NULL"] <- NA


brfss <- brfss %>% rename("AgeGroup" = "Stratification1",
                                "AgeGroupID" = "StratificationID1",
                                "Gender_RaceID" = "StratificationID2") # new_name = old_name

head(brfss)

# Rearrange columns
brfss <- brfss %>% select(YearStart, YearEnd, LocationID, LocationAbbr, LocationDesc,
                     ClassID, Class, TopicID, Topic, AgeGroupID, AgeGroup, Gender_RaceID, Race_Ethnicity, 
                     Gender, QuestionID, Question, Data_Value_Type, Data_Value, 
                     Data_Value_Alt, Data_Value_Footnote, Geolocation)

unique(brfss$Race_Ethnicity)


# Create df of each character column with their ID and distinct value 
# Maybe try to put this into table form 
class <- brfss %>% select(ClassID, Class) %>% distinct()
location <- brfss %>% select(LocationID, LocationAbbr, LocationDesc) %>% distinct()
topic <- brfss %>% select(TopicID, Topic) %>% distinct()
age <- brfss %>% select(AgeGroupID, AgeGroup) %>% distinct()
gender_race <- brfss %>% select(Gender_RaceID, Gender, Race_Ethnicity) %>% distinct()
question <- brfss %>% select(QuestionID, Question, Data_Value_Type) %>% distinct()

# Trends of topics within Overall Health Class
brfss %>% filter(LocationAbbr == "US" & 
                      AgeGroupID == "AGE_OVERALL" & 
                      Gender_RaceID == "OVERALL" &
                   Data_Value_Type == "Percentage" &
                   Class == "Overall Health") %>% 
  group_by(Class) %>% ggplot(aes(x = YearStart, y = Data_Value, color = Question)) + 
  geom_line() +
  theme_minimal() + 
  guides(color = guide_legend(ncol = 1)) +
  scale_y_log10() +
  ggtitle("Trends in Overall Health") + 
  xlab("Year") +
  ylab("Response Percent") +
  theme(legend.position = "bottom",
        axis.title.x = element_text(color="black",size=10),
        axis.title.y = element_text(color="black", size=10),
        axis.text.x = element_text(size=8, angle = 60, vjust = 1, hjust = 1),
        axis.text.y = element_text(size=10),
        plot.title = element_text(color = "black", size=12))

# Trends in questions that have a mean value type 
brfss %>% filter(LocationAbbr == "US" & 
                      AgeGroupID == "AGE_OVERALL" & 
                      Gender_RaceID == "OVERALL" &
                      Data_Value_Type == "Mean") %>% 
  group_by(Class) %>% ggplot(aes(x = YearStart, 
                                 y = Data_Value, 
                                 linetype = Question,
                                 color = Question)) + 
  geom_line() + 
  facet_grid(~Class) + 
  theme_minimal() + 
  guides(color = guide_legend(ncol = 1),
         linetype = guide_legend(ncol = 1)) +
  ggtitle("Trends in Overall Health") + 
  xlab("Year") +
  ylab("Avg No. of days") +
  theme(legend.position = "bottom",
        axis.title.x = element_text(color="black",size=10),
        axis.title.y = element_text(color="black", size=10),
        axis.text.x = element_text(size=8, angle = 60, vjust = 1, hjust = 1),
        axis.text.y = element_text(size=10),
        plot.title = element_text(color = "black", size=12))

# Trends in overall health between genders

unique(brfss$Gender_RaceID)

brfss %>% filter((Gender_RaceID == "FEMALE" | Gender_RaceID == "MALE") &
                      Class == "Overall Health" & LocationAbbr == "US" & 
                      AgeGroupID == "AGE_OVERALL" &
                      Data_Value_Type == "Percentage") %>% 
  ggplot(aes(x = YearEnd, y = Data_Value, color = Question, linetype = Question)) + 
  geom_line() +
  facet_grid(~ Gender_RaceID) + 
  theme_minimal() +
  guides(color = guide_legend(ncol = 1)) + 
  ggtitle("Trends in Overall Health By Gender") + 
  xlab("Year") +
  ylab("Response Percent") +
  theme(legend.position = "bottom",
        axis.title.x = element_text(color="black",size=10),
        axis.title.y = element_text(color="black", size=10),
        axis.text.x = element_text(size=8, angle = 60, vjust = 1, hjust = 1),
        axis.text.y = element_text(size=10),
        plot.title = element_text(color = "black", size=12))

# Trends in overall health between races

brfss %>% filter((Gender_RaceID == "HIS" | Gender_RaceID == "NAA" | 
                       Gender_RaceID == "ASN" | Gender_RaceID == "BLK" |
                       Gender_RaceID == "WHT") &
                      Class == "Overall Health" & LocationAbbr == "US" & 
                      AgeGroupID == "AGE_OVERALL" &
                      Data_Value_Type == "Percentage") %>% 
  ggplot(aes(x = YearEnd, y = Data_Value, color = Question, linetype = Question)) + 
  geom_line() +
  facet_grid(~ Gender_RaceID) + 
  theme_minimal() +
  guides(color = guide_legend(ncol = 1)) +
  ggtitle("Trends in Overall Health by Race") + 
  xlab("Year") +
  ylab("Response Percent") +
  theme(legend.position = "bottom",
        axis.title.x = element_text(color="black",size=10),
        axis.title.y = element_text(color="black", size=10),
        axis.text.x = element_text(size=8, angle = 60, vjust = 1, hjust = 1),
        axis.text.y = element_text(size=10),
        plot.title = element_text(color = "black", size=12))
```

```{r}
# Trends in mental health and cognitive decline
brfss %>% filter(Gender_RaceID == "OVERALL" &
                      (Class == "Cognitive Decline" | Class == "Mental Health") &
                      LocationAbbr == "US" & 
                      AgeGroupID == "AGE_OVERALL" &
                      Data_Value_Type == "Percentage") %>% 
  ggplot(aes(x = YearEnd, y = Data_Value, color = Question, linetype = Class)) + 
  geom_line(linewidth = 0.75) +
  facet_grid( ~QuestionID) + 
  theme_minimal() +
  guides(color = guide_legend(ncol = 1)) +
  ggtitle("Trends in Cognitive and Mental Health") + 
  xlab("Year") +
  ylab("Response Percent") +
  theme(legend.position = "bottom",
        axis.title.x = element_text(color="black",size=10),
        axis.title.y = element_text(color="black", size=10),
        axis.text.x = element_text(size=8, angle = 60, vjust = 1, hjust = 1),
        axis.text.y = element_text(size=10),
        plot.title = element_text(color = "black", size=12))


brfss %>% filter((Gender_RaceID == "FEMALE" | Gender_RaceID == "MALE") &
                      (Class == "Cognitive Decline" | Class == "Mental Health") &
                      LocationAbbr == "US" & 
                      AgeGroupID == "AGE_OVERALL" &
                      Data_Value_Type == "Percentage") %>% 
  ggplot(aes(x = YearEnd, y = Data_Value, color = Question)) + 
  geom_line(linewidth = 0.75) +
  facet_grid( ~Gender_RaceID) + 
  theme_minimal() + 
  guides(color = guide_legend(ncol = 1)) +
  ggtitle("Trends in Cognitive and Mental Health By Gender") + 
  xlab("Year") +
  ylab("Response Percent") +
  theme(legend.position = "bottom",
        axis.title.x = element_text(color="black",size=10),
        axis.title.y = element_text(color="black", size=10),
        axis.text.x = element_text(size=8, angle = 60, vjust = 1, hjust = 1),
        axis.text.y = element_text(size=10),
        plot.title = element_text(color = "black", size=12))

brfss %>% filter((Gender_RaceID == "HIS" | Gender_RaceID == "NAA" | 
                       Gender_RaceID == "ASN" | Gender_RaceID == "BLK" |
                       Gender_RaceID == "WHT") &
                      (Class == "Cognitive Decline" | Class == "Mental Health") &
                      LocationAbbr == "US" & 
                      AgeGroupID == "AGE_OVERALL" &
                      Data_Value_Type == "Percentage") %>% 
  ggplot(aes(x = YearEnd, y = Data_Value, color = Question)) + 
  geom_line(linewidth = 0.75) +
  facet_grid( ~Gender_RaceID) + 
  theme_minimal() + 
  guides(color = guide_legend(ncol = 1)) +
  ggtitle("Trends in Cognitive and Mental Health By Race") + 
  xlab("Year") +
  ylab("Response Percent") +
  theme(legend.position = "bottom",
        axis.title.x = element_text(color="black",size=10),
        axis.title.y = element_text(color="black", size=10),
        axis.text.x = element_text(size=8, angle = 60, vjust = 1, hjust = 1),
        axis.text.y = element_text(size=10),
        plot.title = element_text(color = "black", size=12))

```


```{r}
# Created a function for the rest of the plots
class_plot <- function(class_name){
  
  overall_plot <- brfss %>% filter(Gender_RaceID == "OVERALL" &
                     Class == class_name &
                     LocationAbbr == "US" &
                     AgeGroupID == "AGE_OVERALL" &
                     Data_Value_Type == "Percentage") %>% 
    ggplot(aes(x = YearEnd, y = Data_Value, color = Question)) + 
    geom_line() +
    facet_grid( ~QuestionID) + 
    theme_light() + 
    theme(legend.position = "bottom") +
    guides(color = guide_legend(ncol = 1))


  gender_plot <- brfss %>% filter((Gender_RaceID == "FEMALE" | Gender_RaceID == "MALE") &
                     Class == class_name &
                     LocationAbbr == "US" & 
                     AgeGroupID == "AGE_OVERALL" &
                     Data_Value_Type == "Percentage") %>% 
    ggplot(aes(x = YearEnd, y = Data_Value, color = Question)) + 
    geom_line() +
    facet_grid( ~Gender_RaceID) + 
    theme_light() + 
    theme(legend.position = "bottom") +
    guides(color = guide_legend(ncol = 1))

  race_plot <- brfss %>% filter((Gender_RaceID == "HIS" | Gender_RaceID == "NAA" | 
                      Gender_RaceID == "ASN" | Gender_RaceID == "BLK" |
                      Gender_RaceID == "WHT") &
                        Class == class_name &
                        LocationAbbr == "US" &
                        AgeGroupID == "AGE_OVERALL" &
                        Data_Value_Type == "Percentage") %>% 
    ggplot(aes(x = YearEnd, y = Data_Value, color = Question)) + 
    geom_line() +
    facet_grid( ~Gender_RaceID) + 
    theme_light() + 
    theme(legend.position = "bottom") +
    guides(color = guide_legend(ncol = 1))
  
  list(overall_plot, gender_plot, race_plot)
}

```

```{r}
# Run the function to create the list 
plot_list = list()

for (i in unique(brfss$Class)[c(2,4,5,6)]){
  plot_list[[i]] <- class_plot(i)
}
```

```{r}
unique(brfss$Class)[c(2,4,5,6)]
```

**Uncomment below sections to see graphs**
```{r}
# plot_list[["Nutrition/Physical Activity/Obesity"]]
```

```{r}
# plot_list[["Screenings and Vaccines"]]
```

```{r}
# plot_list[["Smoking and Alcohol Use"]]
```

```{r}
# plot_list[["Caregiving"]]
```

### STATISTICAL ANALLYSIS

Mortality rate of Alzheimer's across the state
```{r}
# https://www.cdc.gov/nchs/pressroom/sosmap/alzheimers_mortality/alzheimers_disease.htm

# https://www.alz.org/alzheimers-dementia/facts-figures

alz_state <- read_csv("https://raw.githubusercontent.com/D-hartog/DATA607/main/FINAL_PROJ/Alzheimer's_mortality_rates.csv")

alz_state$YEAR <- as.numeric(alz_state$YEAR)

alz_state <-  alz_state %>% filter(YEAR == 2021)

unique(population$Location)[c(-10, - 1)]

alz_state$Location <- unique(population$Location)[c(-10, - 1)]


states <- map_data("state")
states$region <- str_to_title(states$region)

states <- states[order(states$order),]

alz_state %>% 
  merge(states, by.x = "Location", by.y = "region") %>%
  arrange(group, order) %>% 
  ggplot(aes(x = long, y = lat)) + 
  geom_polygon(aes(group = group, fill = RATE), color = "white", linewidth = 0.1) + 
  coord_map() +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  theme_minimal() + 
  ggtitle("Alzheimer's Mortality Rate (2021)") + 
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        plot.title = element_text(color = "black", size=24))

```

```{r}
# ALZHEIMERS MORTALITY BY STATE
#
alz_data <- read_csv("https://raw.githubusercontent.com/D-hartog/DATA607/main/FINAL_PROJ/Alzheimer's_mortality_rates.csv")

head(alz_data)
# https://www.alz.org/alzheimers-dementia/facts-figures

`%nin%` <-  Negate(`%in%`)

sub_cog <- brfss %>% filter(QuestionID == "Q30", YearStart == 2021, AgeGroupID == "AGE_OVERALL", 
                 Gender_RaceID == "OVERALL", LocationAbbr %nin% c("MDW", "NRE", "US", "WEST", "SOU"))

head(sub_cog)

alz_data %>% filter(YEAR == 2021) %>% select(YEAR, STATE, RATE) %>% 
  merge(sub_cog, by.x = "STATE", by.y = "LocationAbbr") %>%
  ggplot(aes(x = Data_Value, y = RATE)) +
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE, color = "darkturquoise")

alz_data %>% filter(YEAR == 2021) %>% select(YEAR, STATE, RATE) %>% 
  merge(sub_cog, by.x = "STATE", by.y = "LocationAbbr") %>% 
  summarise(Correlation = cor(Data_Value, RATE, use = "complete.obs"))

sub_cog2 <- brfss %>% filter((QuestionID == "Q30" | QuestionID == "Q31" | QuestionID == "Q41" | QuestionID == "Q42") &
                            YearStart == 2021, AgeGroupID == "AGE_OVERALL", 
                            Gender_RaceID == "OVERALL", 
                            LocationAbbr %nin% c("MDW", "NRE", "US", "WEST", "SOU"))

sub_cog2 <- sub_cog2 %>% pivot_wider(id_cols = LocationAbbr, names_from = QuestionID, values_from = Data_Value)

new_data <- alz_data %>% filter(YEAR == 2021) %>% select(YEAR, STATE, RATE) %>% 
  merge(sub_cog2, by.x = "STATE", by.y = "LocationAbbr")

sub_cog2_reg <- lm(RATE ~ Q30 + Q31 + Q41 + Q42, data = new_data)
summary(sub_cog2_reg)
```

# CONCLUSIONS 

**Observations from the data:**

+ There are clear trends in the growth of the aging population with over 25% of the US population in 2022 being 55 or older. We can probably expect that number to continue to rise.

+ Trends the mortality rate in the US spiked during COVID and continued to rise in 2021. 

+ COVID-19 and Chronic liver disease over took Strokes and the Flu in the top 10 leading causes of death in the US during 2020 and 2021. 

+ Since 1999 there appears to be a steady climb in the rates of deaths from Alzheimer’s and Unintentional injuries. 

+ Trends in mortality for age groups over 65 stayed relatively unchanged from 2021 to 2022. 

+ Alzheimer's Disease was not a top 10 leading cause of death in the age group 65 - 74 years old but historically it has been in the top 10 leading causes of death across the US. (however it was the 11th leading cause)

+ Regarding overall health, 60% of respondents report that they have high blood pressure and are on blood pressure medications. 

+ Over 50% of the respondents in the BRFSS survey reported that their health is “good”, “very good”, “excellent”

+ Only 50% of respondents who reported subjective cognitive decline or memory loss have spoken to a medical professional about it.

+ 40-60% of Black, Hispanic, and Native American/Alaskan Natives reported that subjective cognitive decline has interfered with their social and daily activities; Compare with 25-30% of White respondents. 

+ Very weak to correlations exist between a states Alzheimer's death rate and  the percentage of individuals who reported mild cognitive decline. 

# LIMITATIONS AND FUTHER CONSIDERATIONS

**Limitations** 
+ Data is only up to 2021/2022

+ Trends in US mortality rates overtime are age adjusted - useful for comparing to other populations 

+ Limited data from all states to generate any significant correlations between Alzheimer’s death rates and subjective reports of cognitive decline. 

+ Many factors relate to cognitive decline and the onset of Alzheimer’s and responses on the BFRSS might not be a good indicator of mortality 

**Future directions**

+ Since we have data on age adjusted death rates comparisons between states and other countries might be interesting 

+ Investigate trends from the BFRSS among the different age groups

+ Comparing mortality data to other countries might shed light on effective strategies in other countries. 

